\Large \bf DRAFT: Direct-style process creation on Linux

### 2021-04-08 research/paper/paper.tex
paper

\usepackage{tikz}
\usepackage{dblfloatfix}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{pifont}
\usepackage{comment}

@@

\section{Introduction}\label{introduction}
Existing process creation interfaces can be divided into three categories:
Fork-style, spawn-style, and direct-style.
\begin{table*}
\begin{tabular}{l|l|l|l}
% BEGIN RECEIVE ORGTBL compare
 & fork & spawn & direct-style\\
\hline
Requirements on parent process & \tbad Single-threaded, small memory & \tgood None & \tgood None\\
Programming model & \tbad Complex (returns twice) & \tgood Simple (single call) & \tgood Simple (imperative)\\
Maximally powerful & \tgood Yes, can call any syscall & \tbad No, limited interface & \tgood Yes, can call any syscall\\
Reporting of results & \tbad Requires IPC & \tbad Not fine-grained & \tgood From individual child syscalls\\
Non-inherited attributes & Mutated by code in child & Set by arguments & Mutated by code in parent\\
% END RECEIVE ORGTBL compare
\end{tabular}
\caption{Features of fork-style vs spawn-style vs direct-style}
\label{tab:styles}
\end{table*}
\begin{comment}
#+ORGTBL: SEND compare orgtbl-to-latex :splice t
|                                | fork                                | spawn                       | direct-style                          |
|--------------------------------+-------------------------------------+-----------------------------+---------------------------------------|
| Requirements on parent process | \tbad Single-threaded, small memory | \tgood None                 | \tgood None                           |
| Programming model              | \tbad Complex (returns twice)       | \tgood Simple (single call) | \tgood Simple (imperative)            |
| Maximally powerful             | \tgood Yes, can call any syscall    | \tbad No, limited interface | \tgood Yes, can call any syscall      |
| Reporting of results           | \tbad Requires IPC                  | \tbad Not fine-grained      | \tgood From individual child syscalls |
| Non-inherited attributes       | Mutated by code in child            | Set by arguments            | Mutated by code in parent             |
% $
\end{comment}
Table \ref{tab:styles} summarizes the differences between each style.

@@

We'll discuss each style in more detail in section \ref{background}.

To perform direct-style process creation on Linux,
we need an API where we explicitly specify in which process we want to make a syscall.
This differs from the normal mode of operation on Linux,
where programs making syscalls implicitly operate on the current process.

Our contribution in this paper is this API, through the \texttt{rsyscall} library.
The \texttt{rsyscall} project develops
language-specific, object-capability-model, low-abstraction libraries for the Linux syscall interface,
bypassing libc.
\texttt{rsyscall} is currently focused on supporting Python,
but can be ported to other languages;
the examples we show in this paper will be in Python,
but generalize easily to other languages.
\texttt{rsyscall} is open source and available from
[Github url].

To perform direct-style process creation,
we use \texttt{rsyscall} to call \texttt{clone} on the ``local'' process to create a new child process.
We call various syscalls in the child process to mutate it until it reaches the desired state,
at which point we can call \texttt{exec} in the child process to start it running on its own.
After that, the process is no longer under our control, and it functions like any other child process.
The child process can be monitored using normal Linux child monitoring syscalls,
such as \texttt{waitid}.

These child process capabilities are created and managed in userspace,
by launching new processes running a syscall server,
which receives syscall requests, performs the syscall, and returns the response.
No kernel modifications are required, and \texttt{rsyscall} is immediately deployable on recent Linux kernel.
We discuss the implementation in depth in section \ref{implementation},
and examine several difficult process-creation related details.

\lstinputlisting[float=h,language=Python,label={basic},caption={Create new process, change CWD, and exec}]
{examples/basic/direct.py}
Listing \ref{basic} shows a program using \texttt{rsyscall}.
We create a new child process using \texttt{clone} on the local process \verb|local|,
and receive back a capability to control it.
We call \texttt{chdir} in the child process to change its working directory to \texttt{/dev}.
Then we call \texttt{execv} in the child,
passing the path of the \texttt{cat} executable
and the arguments to the new executable.
The name of the executable is passed as the first argument,
as is typical in Unix.
Calling \texttt{execv} consumes the capability,
and later use of this child process capability will fail with an exception.

The behavior of system calls made on a remote process through \texttt{rsyscall}
is identical to the behavior of system calls made implicitly on the current process through libc.
This is an API that systems programmers are already familiar with,
and which is already thoroughly documented through Linux system call manpages.
The novelty is constrained to the newly explicit process capabilities.
Besides flattening the learning curve,
this also ensures that direct-style \texttt{clone} will always be at least as powerful as \texttt{fork}.

We show more examples of direct-style process creation in section \ref{examples},
including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.

Direct-style \texttt{clone} has acceptable performance cost,
and can outperform \texttt{fork} in some cases.
The performance overhead of creating processes with direct-style clone
is comparable to the performance cost of creating processes while inside a Linux namespace.
At \twosigma, we have deployed direct-style process creation in production
inside a library for distributed system deployment, Toplevel.
We evaluate in further detail in section \ref{evaluation}.
\section{Background on process-creation}\label{background}
As mentioned in the introduction,
on most operating systems today,
the available interfaces for process creation
are either "fork-style" or "spawn-style".
A few operating systems have a third style, which we refer to as "direct-style".

\subsection{Fork-style}
By far the most well-known fork-style interface is Unix's fork.\cite{forkhist}
With a fork-style interface,
a process, called the parent process, calls \texttt{fork},
and a copy of that process is made,
called the child process,
differing only in a few attributes such as process id.
The same program continues executing in both processes,
with a different return value from fork to allow it to distinguish whether it is running in the parent or the child.
Typically, the program will branch based on the return value
and, if running in the child,
call various system calls to modify the state of the new process.

@@

However, we believe that a natural port of direct-style process creation to Linux
provides an interface that is just as simple to use as fork-style or spawn-style on Linux,
without their disadvantages.
Such an interface, like fork-style and spawn-style,
would implicitly copy some attributes to the new process,
rather than being capability-oriented and requiring all attributes to be specified explicitly.
This is the kind of direct-style interface that we contribute in this paper.
\section{Examples}\label{examples}
In this section,
we'll demonstrate direct-style process creation
by demonstrating several advanced uses of processes
using more sophisticated Linux features.
For concision, we assume in these examples that we are running with sufficient permissions in these examples,
but these examples can all be run without privileges with appropriate use of \verb|CLONE_NEWUSER|.

@@

\lstinputlisting[float=*ht,language=Python,label={fds},caption={Passing down FDs}]
{examples/fds/direct.py}
In Listing \ref{fds},
we first open a file with read-write permission in the local process.
Then we create a child process,
which inherits all file descriptors from its parent.
We indicate that we want to operate on the child's inherited copy of the file descriptor with \verb|child.inherit_fd|.
Then we disable \texttt{CLOEXEC} on the child's file descriptor,
which on Linux we can do by clearing the FD flags with \verb|fcntl(fd, F_SETFD, 0)|.
Finally, we execute some executable in the child process,
passing the file descriptor number as an argument;
since we disabled \texttt{CLOEXEC} on the file descriptor,
the executable will be able to use that file descriptor through that file descriptor number.
\subsection{Non-shared-memory concurrency}
Processes run concurrently,
which allows exploiting the parallelism of the underlying hardware.
Since processes don't share memory,
they can provide a less complex parallel programming environment
than shared-memory thread-based approaches.

@@

The use of namespaces also has a significant impact,
both when the parent process is in a namespace, and when the child process is created in a namespace.
We perform the benchmark with no clone flags, and with the common clone flags
\verb|CLONE_NEWNS| and \verb|CLONE_NEWPID|, first on their own and then together.
This benchmark essentially measures the performance of the Linux namespaces system, not the work in this paper,
but it is illustrative to compare it to the performance of \texttt{rsyscall}.
Note that the cost of creating a new child process from a parent process that is already inside a namespace
(the difference between the \texttt{clone()} and \texttt{clone(NS|PID)} parent processes)
is around 2 milliseconds, comparable to the cost of running with \texttt{rsyscall}.
Note also that the overhead of creating a new process inside a new namespace
(the difference between the \texttt{clone()} and \texttt{clone(NS|PID)} child processes)
is around 15 milliseconds; subtantially greater than the overhead of \texttt{rsyscall}.
\subsection{Usage in the real world}\label{realworld}
At \twosigma we have used \texttt{rsyscall} and direct-style process creation
to implement an internal distributed system deployment library, written in Python,
which we'll refer to here as Toplevel.
Toplevel is in use in production,
and is extensively used as part of testing and development.
Toplevel consists of a collection of functions and modules
for each component in our system,
which start up components using direct-style process creation,
all from a single Python parent process.

One of the major benefits of direct-style process creation has been easy use of file descriptor inheritance.
Our experience has been that with fork-style or spawn-style interfaces,
all file descriptor logic must be centralized,
whereas file descriptors in a new process can be built up over time using direct-style,
simplifying the implementation.
We've found that even programmers without substantial prior experience with Unix programming
are able to use file descriptor passing as a normal feature of their development process with relative ease,
writing code to support new programs and passing down file descriptors to those programs without issue.

Easy file descriptor passing has, in turn, allowed us to heavily use socket activation techniques
to start up services in parallel,
substantially speeding up system startup.\cite{socketactivation}
We are able to bootstrap connections over our internal shared memory transports
over pre-created, passed-down file descriptors,
and our management interfaces listen on passed-down sockets;
this allows starting up services in parallel, rather than in dependency order.
Benchmarking one test built with an old library and its equivalent built with Toplevel,
the old test took an average of 221 seconds to run,
mostly spent in system startup,
while the new test takes an average of 16 seconds.

Easy file descriptor inheritance has also allowed us to treat file descriptors as a uniform interface in Toplevel.
We pass around file descriptors freely inside Toplevel.
A given component can implemented with a separate process which has the file descriptor passed down,
running a program written in native code for performance,
or implemented in-process with a Python implementation for flexibility;
either is possible behind a uniform file descriptor interface.

We've also been able to create complicated pipelines in Toplevel,
as in section \ref{pipe}.

We've also made use of namespaces,
chiefly user namespaces and pid namespaces.
For legacy applications which directly start subprocesses,
we prevent process leaking with a pid namespace.
This is part of strategy of noninterference and cleanup
that ensures that multiple users of Toplevel
can coexist simultaneously on the same host,
even if running the same system.

Toplevel also makes uses of other features of \texttt{rsyscall} which are not covered in this paper.
These include the ability to operate on remote hosts through capabilities for remote processes,
and the ability to perform system calls in parallel across a pool of processes.
These have significant synergy with direct-style process creation
in forming an overall powerful system.

>>>>

\usepackage{tikz}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{pifont}
\usepackage{comment}

@@

\section{Introduction}\label{introduction}
Existing process creation interfaces can be divided into three categories:
Fork-style, spawn-style, and direct-style.

@@

Our contribution is an implementation of direct-style process creation for Linux,
making use of our work on the \texttt{rsyscall} project.
The \texttt{rsyscall} project develops language-specific ``process-independent'' system call libraries for Linux,
which replace libc system call wrappers,
and allow a program to perform cross-process system calls.
\texttt{rsyscall} works entirely in userspace
and can be deployed on existing Linux systems without kernel modifications.
\texttt{rsyscall} is currently focused on supporting, and is implemented in, Python,
but can be ported to other languages;
the examples we show in this paper will be in Python,
but generalize easily to other languages.

Our direct-style process creation implementation uses \texttt{clone} to create a child process in an embryonic state
which can then be manipulated from user code in the parent process using syscalls with \texttt{rsyscall}.
A direct-style \texttt{clone} has all the advantages of direct-style process creation
over Linux's fork-style \texttt{fork} and spawn-style \texttt{posix\_spawn}:
\begin{compactitem}
\item
  Processes can be created correctly and performantly from any parent process,
  with a straightforward, familiar API.
\item
  Processes can be created with the full set of Linux features.
\item
  Syscall results are reported individually to the parent process.
  % rather than batched together as with spawn-style or manually reported via IPC as with fork-style.
\end{compactitem}

We've implemented a variety of useful applications using direct-style process creation on Linux,
which we'll demonstrate in this paper.
Direct-style \texttt{clone} has minimal performance cost in normal situations,
and can outperform \texttt{fork} in some cases.
We have deployed direct-style process creation in production internally at \twosigma.

Our implementation is distributed as a feature of \texttt{rsyscall},
and is immediately deployable, entirely in userspace, on recent Linux kernels.
\texttt{rsyscall} is open source and available from
[Github url].
% \url{https://github.com/catern/rsyscall}.

This paper is organized in several sections.
\begin{compactitem}
\item Section \ref{background} provides background on fork-style, spawn-style, and direct-style process creation,
and considers their advantages and disadvantages in more depth.
\item Section \ref{overview} discusses the design of the API,
including our usage of \texttt{rsyscall}.
\item Section \ref{examples} shows examples of direct-style process creation in Python on Linux,
including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
\item Section \ref{implementation} explores our implementation in depth,
and discusses several difficult process-creation related details of \texttt{rsyscall}.'s implementation.
\item Section \ref{evaluation} evaluates our results using both quantitative and qualitative approaches.
\item Section \ref{related_work} examines other work which also touches on the topics covered by this paper.
\item Section \ref{future_work} lists several directions for future work
related to direct-style process creation and to \texttt{rsyscall}.
\item Section \ref{conclusions} briefly outlines our conclusions about direct-style and the future of process creation.
\end{compactitem}
\section{Background on process-creation}\label{background}
As mentioned in the introduction,
on most operating systems today,
the available interfaces for process creation
are either "fork-style" or "spawn-style".
A few operating systems have a third style, which we refer to as "direct-style".

\begin{table*}
\begin{tabular}{l|l|l|l}
% BEGIN RECEIVE ORGTBL compare
 & fork-style & spawn-style & direct-style\\
\hline
Works with any caller & \tbad No threads or large memory & \tgood Yes & \tgood Yes\\
Programming model & \tbad Complex (returns twice) & \tgood Simple (single call) & \tgood Simple (imperative)\\
Maximally powerful & \tgood Yes, can call any syscall & \tbad No, limited interface & \tgood Can call any syscall\\
Error handling & \tbad Poor, requires IPC & \tbad Poor, not fine-grained & \tgood Reports individual errors\\
Non-copied attributes & Mutated by code in child & Set by arguments & Mutated by code in parent\\
% END RECEIVE ORGTBL compare
\end{tabular}
\caption{Features of fork-style vs spawn-style vs direct-style}
\label{tab:ease}
\end{table*}
\begin{comment}
#+ORGTBL: SEND compare orgtbl-to-latex :splice t
|                       | fork-style                       | spawn-style                  | direct-style                     |
|-----------------------+----------------------------------+------------------------------+----------------------------------|
| Works with any caller | \tbad No threads or large memory | \tgood Yes                   | \tgood Yes                       |
| Programming model     | \tbad Complex (returns twice)    | \tgood Simple (single call)  | \tgood Simple (imperative)       |
| Maximally powerful    | \tgood Yes, can call any syscall | \tbad No, limited interface  | \tgood Can call any syscall      |
| Error handling        | \tbad Poor, requires IPC         | \tbad Poor, not fine-grained | \tgood Reports individual errors |
| Non-copied attributes | Mutated by code in child         | Set by arguments             | Mutated by code in parent        |
% $
\end{comment}
\subsection{Fork-style}
By far the most well-known fork-style interface is Unix's fork.\cite{forkhist}
With a fork-style interface,
a process, called the parent process, calls \texttt{fork},
and a copy of that process is made,
called the child process,
differing only in a few attributes such as process id.
The same program continues executing in both processes,
with a different return value from fork to allow it to distinguish whether it is running in the parent or the child.
Typically, the program will branch based on the return value
and, if running in the child,
call various system calls to modify the state of the new process.

@@

However, we believe that a natural port of direct-style process creation to Linux
provides an interface that is just as simple to use as fork-style or spawn-style on Linux,
without their disadvantages.
Such an interface, like fork-style and spawn-style,
would implicitly copy some attributes to the new process,
rather than being capability-oriented and requiring all attributes to be specified explicitly.
This is the kind of direct-style interface that we contribute in this paper.
\section{Overview of API}\label{overview}
In most operating system APIs, including the ones available on Linux,
when a program makes a syscall, it implicitly operates on the current process.
To perform direct-style process creation on Linux,
we need an API where we instead explicitly specify in which process we want to make a syscall.

\texttt{rsyscall} provides this API.
The \texttt{rsyscall} project develops
language-specific, object-capability-model, low-abstraction libraries for the Linux syscall interface,
bypassing libc.

To perform direct-style process creation,
we use \texttt{rsyscall} to call \texttt{clone} on the ``local'' process to create a new child process.
We call various syscalls in the child process to mutate it until it reaches the desired state,
at which point we can call \texttt{execve} in the child process to start it running on its own.
After that, the process is no longer under our control, and it functions like any other child process.
The child process can be monitored using normal Linux child monitoring syscalls,
such as \texttt{waitid}.

\lstinputlisting[float=ht,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
{examples/basic/direct.py}
Listing \ref{basic} shows this in action in Python.
We create a new child process using \texttt{clone} on the local process \verb|local|,
and receive back a capability to control it.
We call \texttt{chdir} in the child process to change its working directory to \texttt{/dev}.
Then we call \texttt{execve} in the child,
passing the path of the \texttt{cat} executable,
the arguments to the new executable,
and the unchanged set of environment variables.
The name of the executable is passed as the first argument,
as is typical in Unix.
Calling \texttt{execve} consumes the capability,
and later use of this child process capability will fail with an exception.

The behavior of system calls made on a remote process through \texttt{rsyscall}
is identical to the behavior of system calls made implicitly on the current process through libc.
This is an API that systems programmers are already familiar with,
and which is already thoroughly documented through Linux system call manpages.
The novelty is constrained to the newly explicit process capabilities.

The lack of novelty in the API is important not only for the learning curve,
but also for power.
Since we are using conventional Linux system calls,
just in a cross-process way,
direct-style \texttt{clone} will always be at least as powerful as \texttt{fork}.
\section{Examples}\label{examples}
In this section,
we'll demonstrate direct-style process creation
by demonstrating several advanced uses of processes
using more sophisticated Linux features.
For concision, we assume in these examples that we are running with sufficient permissions in these examples,
but these examples can all be run without privileges with appropriate use of \verb|CLONE_NEWUSER|.

@@

\lstinputlisting[float=*ht,language=Python,label={fds},caption={Passing down FDs}]
{examples/fds/direct.py}
In Listing \ref{fds},
we first open a file with read-write permission in the local process.
Then we create a child process,
which inherits all file descriptors from its parent.
We indicate that we want to operate on the child's inherited copy of the file descriptor with \verb|child.inherit_fd|.
Then we disable \texttt{CLOEXEC} on the child's file descriptor,
which on Linux we can do by clearing the FD flags with \verb|fcntl(fd, F_SETFD, 0)|.
Finally, we execute some executable in the child process,
passing the file descriptor number as an argument;
since we disabled \texttt{CLOEXEC} on the file descriptor,
the executable will be able to use that file descriptor through that file descriptor number.
Unlike in listing \ref{basic}, we call the \texttt{execv} wrapper rather than \texttt{execve},
so that we don't need to pass the (unmodified) environment to every \texttt{exec} call in these examples.
\subsection{Non-shared-memory concurrency}
Processes run concurrently,
which allows exploiting the parallelism of the underlying hardware.
Since processes don't share memory,
they can provide a less complex parallel programming environment
than shared-memory thread-based approaches.

@@

The use of namespaces also has a significant impact,
both when the parent process is in a namespace, and when the child process is created in a namespace.
We perform the benchmark with no clone flags, and with the common clone flags
\verb|CLONE_NEWNS| and \verb|CLONE_NEWPID|, first on their own and then together.
This benchmark essentially measures the performance of the Linux namespaces system, not the work in this paper,
but it is illustrative to compare it to the performance of \texttt{rsyscall}.
Note that the cost of creating a new child process from a parent process that is already inside a namespace
(the difference between the \texttt{clone()} and \texttt{clone(NS|PID)} parent processes)
is around 2 milliseconds, comparable to the cost of running with \texttt{rsyscall}.
Note also that the overhead of creating a new process inside a new namespace
(the difference between the \texttt{clone()} and \texttt{clone(NS|PID)} child processes)
is around 15 milliseconds; subtantially greater than the overhead of \texttt{rsyscall}.
\subsection{Usage in the real world}\label{realworld}
At \twosigma we have used \texttt{rsyscall} and direct-style process creation
as a component of an internal distributed system deployment library, written in Python.
This library is in use in production,
and is extensively used as part of testing and development.

Each service deployed by this library is started with custom code,
written using direct-style process creation.
This allows us to pass down file descriptors and initialize process attributes
according to each program's needs;
which in turn has let us make optimizations which previously required too much program-specific customization.

For example,
we are able to bootstrap connections over our internal shared memory transports over passed-down file descriptors,
and our management interfaces listen on passed-down sockets.
This allows substantial parallelism in the startup of services.\cite{socketactivation}
Taking one representative example,
one test built with an old library took an average of 221 seconds to run,
mostly spent in system startup time as processes started up serially.
With parallelized system startup, the same test takes an average of 16 seconds.

We've found that programmers
are able to use file descriptor passing as a normal feature of their development process with relative ease.
Programmers without substantial prior experience with Unix programming
are able to write code to support new services
nd pass down file descriptors to the program without encountering significant issues.

We also flexibly use namespaces where appropriate.
For example, for legacy applications which directly start subprocesses,
we prevent process leaking with a pid namespace.
